\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-05 paper.tex
Minor updates

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress has led to a diversity of ARG definitions and representations.
% We show that a synthesis of these views can be achieved by thinking
% in terms of genomes and transmission
% intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
We argue that classical ARG formulations, which assume
that all events can be known and precisely estimated,
are fundamentally unsuited to the realities
of modern, population-scale datasets.
We describe an alternative formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

\section{A diversity of structures}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of four methods run on the 
\cite{kreitman1983nucleotide} dataset, often used as an
ARG inference benchmark. One source of differences between 
the inferences is the underlying model. For instance, 
the ARG in Fig.~\ref{fig-inferred-args}A is inferred by 
\kwarg\ under parsimony, and therefore has many fewer 
ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
is sampled from the coalescent with recombination with an 
appropriate recombination rate parameter. These differences
are well understood.

@@

The differences between ARGs that we wish to draw attention to here 
are more subtle

>>>>

% NEEDS REWRITE
\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress has led to a diversity of ARG definitions and representations.
% We show that a synthesis of these views can be achieved by thinking
% in terms of genomes and transmission
% intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
We argue that classical ARG formulations, which assume
that all events can be known and precisely estimated,
are fundamentally unsuited to the realities
of modern, population-scale datasets.
We describe an alternative formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

\section{A diversity of structures}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring}
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
One source of differences between 
the inferences is the underlying model. For instance, 
the ARG in Fig.~\ref{fig-inferred-args}A is inferred 
under parsimony, and therefore has many fewer 
ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
is sampled from the coalescent with recombination with an 
appropriate recombination rate parameter. Such differences
are well understood.

@@

[WIP] The differences between ARGs that we wish to draw attention to here 
are more subtle
